On the use of 'boilerplate' statistical analysis sections?

### 2021-03-29 manuscript/draft_manuscript.Rmd
Update draft_manuscript.Rmd

An ideal statistical analysis will use appropriate methods to create insights from the data and inform the research questions. Unfortunately many current statistical analyses are far from ideal, with many researchers using the wrong methods, misinterpreting the results, or failing to adequately check their assumptions [@Goodman2008;@Leek2017]. Some researchers take a "mechanistic" approach to statistics, copying the few methods they know regardless of their appropriateness, and then going through the motions of the analysis [@Stark2018].

@@

Definitions of statistical significance featured strongly in Topic 1 (n = 3,775) and Topic 9 (n = 6,104) combined with hypothesis testing for comparing differences between two groups. Topic 1 reflected applications of Student's t-test assuming a 5% level of statistical significance. Topic 9 referenced similar methods combined with multiple thresholds for declaring statistical significance: \*p<0.05, \*\*p<0.01 and \*\*\*p<0.001. Examples of boilerplate text are give in Box 2.

@@

Group-based hypothesis testing was a recurring theme across topics, with text descriptions varying based on method(s) used (Box 3). For example, frequently occurring words in Topic 6 (n = 4,746) including 'mean', 'sd' and 'sem' reflected the use of descriptive statistics for summarising continuous variables. Sections associated with this topic appeared to be expanded versions of Topics 1 and 9, showing evidence of boilerplate text in the form of descriptive statistics followed by hypothesis testing using Student's t-test, Mann Whitney U or one-way analysis of variance. One-way analysis of variance also featured strongly in Topic 4 (n = 10,163), with text citing common methods for performing post-hoc multiple comparisons. Top matching sections across these topics did not include definitions of groups being compared.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown in Figure~X. 
Of the 12,700 included studies, 9,523 had a statistical methods section, 75% (95% CI 74% to 76%).

ftab = flextable(table) %>%
theme_box() %>%
autofit()
ftab

Some of the non-missing statistics sections were only one word, including "ANOVA", "t-test", "SPSS" and even "SSPS". The median length of the section was 129 words with an inter-quartile range of 71 to 219 words.

The clustering algorithm found groups that were purely sample size calculations (topic 2), pilot studies (topic 5), safety/tolerability studies (topic 6) and repeated measures ANOVA (topic 10). There were cases where the exact same method section had been ...

@@

* "Shapiro Wilk test was used as normality test. Continuous variables were compared using Student t-test and Mann-Whitney U test when the data were not normally distributed. Categorical variables were compared using Pearson’s chi-squared test and Fisher’s exact test. Paired data were analyzed using Paired t-test and Wilcoxon signed rank test when data were not normally distributed"
* "Comparisons between categorical variables will be made either using chi square or Fisher exact test. Continuous data will be compared using the Student’s t-test or Mann-Whitney U test. Two sided p values of less than 0.05 will be considered statistically significant."

@@

Many researchers are using lazy practice by copying a standard "boilerplate" statistical methods section, likely cut-and-pasting from other researchers or projects. This is a strong sign of the ritualistic practice of statistics where researchers go through the motions rather than using conscientious practice [@Stark2018]. This is concerning because using the wrong statistical methods can reduce the potential of study, or worse, invalidate the entire study. Poor statistical practice is a key driver of the ongoing reproducibility crisis in science [@Ioannidis2014].

@@

A frequent theme in the boilerplate statistical methods is the definition of statistical significance, nearly always using a p-value at the 5% level. This widespread use of statistical significance is troubling giving the bright-line thinking it engenders [mcshane] and the common misinterpretations of p-values [goodman].

@@

@@

We did not check whether papers used the correct methods, and for some simple studies a 'boilerplate' statistical methods section would be fine.

@@

We only examined one journal and one trial registry and hence our results may not be generalisable to all journals or registries, especially those that consistently use a statistical reviewer.

>>>>

An ideal statistical analysis will use appropriate methods to create insights from the data and inform the research questions. Unfortunately many current statistical analyses are far from ideal, with many researchers using the wrong methods, misinterpreting the results, or failing to adequately check their assumptions [@2008;@Leek2017]. Some researchers take a "mechanistic" approach to statistics, copying the few methods they know regardless of their appropriateness, and then going through the motions of the analysis [@Stark2018].

@@

Definitions of statistical significance featured strongly in Topic~1 (n = 3,775) and Topic 9 (n = 6,104) combined with hypothesis testing for comparing differences between two groups. Topic~1 reflected applications of Student's t-test assuming a 5% level of statistical significance. Topic~9 referenced similar methods combined with multiple thresholds for declaring statistical significance by asterisk: "\*p<0.05, \*\*p<0.01 and \*\*\*p<0.001", a practice that has been criticised [@Wasserstein2019]. Examples of boilerplate text are in Box~2.

@@

Group-based hypothesis testing was a recurring theme across topics, with text descriptions varying based on method(s) used (Box 3). Frequently occurring words in Topic 6 (n = 4,746) including 'mean', 'sd' and 'sem' reflected the use of descriptive statistics for summarising continuous variables. Sections associated with this topic appeared to be expanded versions of Topics 1 and 9, showing evidence of boilerplate text in the form of descriptive statistics followed by hypothesis testing using Student's t-test, Mann Whitney U or one-way analysis of variance. One-way analysis of variance also featured strongly in Topic 4 (n = 10,163), with text citing common methods for performing post-hoc multiple comparisons. Top matching sections across these topics did not include definitions of groups being compared.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown in Figure~X. 
Of the 12,700 included studies, 9,523 (75%) had a statistical methods section.

The median length of the section was 129 words with an inter-quartile range of 71 to 219 words.
Some methods sections were only one word, including "ANOVA", "t-test", "SPSS" and even "SSPS".

The clustering algorithm found groups that were purely sample size calculations (topic 2), pilot studies (topic 5), safety/tolerability studies (topic 6) and repeated measures ANOVA (topic 10). There were cases where the exact same method section had been re-used in a different study.

@@

* "Shapiro Wilk test was used as normality test. Continuous variables were compared using Student t-test and Mann-Whitney U test when the data were not normally distributed. Categorical variables were compared using Pearson’s chi-squared test and Fisher’s exact test. Paired data were analyzed using Paired t-test and Wilcoxon signed rank test when data were not normally distributed."
* "Comparisons between categorical variables will be made either using chi square or Fisher exact test. Continuous data will be compared using the Student’s t-test or Mann-Whitney U test. Two sided p values of less than 0.05 will be considered statistically significant."
___

ftab = flextable(table) %>%
theme_box() %>%
autofit()
ftab
```

@@

@@

A frequent theme in the boilerplate statistical methods is the definition of statistical significance, nearly always using a p-value at the 5% level. This widespread use of statistical significance is troubling giving the bright-line thinking it engenders [@McShane2019] and the common misinterpretations of p-values [@Goodman2008].

@@

Many researchers are using lazy practice by copying a standard "boilerplate" statistical methods section, likely cut-and-pasting from other researchers or projects. This is a strong sign of the ritualistic practice of statistics where researchers go through the motions rather than using conscientious practice [@Stark2018]. This is concerning because using the wrong statistical methods can reduce the value of study, or worse, invalidate the entire study. These mistakes are avoidable and are wasting of thousands of hours of researchers' time and the time of patients and volunteers. Poor statistical practice is a key driver of the ongoing reproducibility crisis in science [@Ioannidis2014].

@@

We did not check whether papers used the correct methods, and for some simple studies a 'boilerplate' statistical methods might be adequate.

@@

We only examined one large journal and one trial registry and hence our results may not be generalisable to all journals or registries, especially those that consistently use a statistical reviewer.
